---
title: "Cyclistic Bike-Share"
author: "Uko Etukho"
date: "1/8/2022"
output: html_document
---




#### INTRODUCTION
Cyclistic Bike-Share case study is a study on the Bike-Share users for a period of 12 months to gain insights into user trends. It is a Google Data Analytics Certificate Capstone project to demonstrate data analytics.

#### BUSINESS TASK
The goal of this case study is to gain insights into different types of user trends and how these trends could help in developing a marketing strategy to persuade casual users to buy into the membership program of the company.

#### DATA SOURCE
The data used in this study is provided for public use by Motivate International Inc. under this [license](https://ride.divvybikes.com/data-license-agreement).

#### DATA CLEANING AND TRANSFORMATION
Analysis and visualizations will be done in the R Studio and Tableau.

##### R STUDIO
Preparing the work environment.
```{r message=FALSE}
library(tidyverse)
library(here)
library(skimr)
library(janitor)
library(lubridate)

```
After preparing the working environment by loading the relevant packages into R Studio, I now loaded the required data sets into the environment.

```{r message=FALSE, results='hide'}
divvy_04 <- read_csv("C:/Users/Dell 5580/Documents/Google Data Analytics/divvy_trips/202004-divvy-tripdata.csv")
glimpse(divvy_04)

```

```{r message=FALSE, results='hide'}
divvy_05 <- read_csv("C:/Users/Dell 5580/Documents/Google Data Analytics/divvy_trips/202005-divvy-tripdata.csv")
spec(divvy_05)
glimpse(divvy_05)
```

```{r message=FALSE, results='hide'}
divvy_06 <- read_csv("C:/Users/Dell 5580/Documents/Google Data Analytics/divvy_trips/202006-divvy-tripdata.csv")
glimpse(divvy_06)

```

```{r message=FALSE, results='hide'}
divvy_07 <- read_csv("C:/Users/Dell 5580/Documents/Google Data Analytics/divvy_trips/202007-divvy-tripdata.csv")
glimpse(divvy_07)

```

```{r message=FALSE, results='hide'}
divvy_08 <- read_csv("C:/Users/Dell 5580/Documents/Google Data Analytics/divvy_trips/202008-divvy-tripdata.csv")
glimpse(divvy_08)

```

```{r message=FALSE, results='hide'}
divvy_09 <- read_csv("C:/Users/Dell 5580/Documents/Google Data Analytics/divvy_trips/202009-divvy-tripdata.csv")
glimpse(divvy_09)

```

```{r message=FALSE, results='hide'}
divvy_10 <- read_csv("C:/Users/Dell 5580/Documents/Google Data Analytics/divvy_trips/202010-divvy-tripdata.csv")
glimpse(divvy_10)

```

```{r message=FALSE, results='hide'}
divvy_11 <- read_csv("C:/Users/Dell 5580/Documents/Google Data Analytics/divvy_trips/202011-divvy-tripdata.csv")
glimpse(divvy_11)

```

```{r message=FALSE, results='hide'}
divvy_12 <- read_csv("C:/Users/Dell 5580/Documents/Google Data Analytics/divvy_trips/202012-divvy-tripdata.csv")
glimpse(divvy_12)

```

```{r message=FALSE, results='hide'}
divvy_q1 <- read_csv("C:/Users/Dell 5580/Documents/Google Data Analytics/divvy_trips/Divvy_Trips_2020_Q1.csv")
glimpse(divvy_q1)

```

Checking the data frames structure for incongruencies.
```{r results='hide'}
str(divvy_04)
str(divvy_05)
str(divvy_06)
str(divvy_07)
str(divvy_08)
str(divvy_09)
str(divvy_10)
str(divvy_11)
str(divvy_12)
str(divvy_q1)

```

After noticing structure incongruencies in divvy_04 and divvy_12 data frames, I changed the 'started_at' and 'ended_at' column types in divvy_04 from character to date-time types and also changed the 'start_station_id' and 'end_station_id' columns types in divvy_12 from character to number types.

```{r, results='hide'}
divvy_04$started_at <-  mdy_hm(divvy_04$started_at)
divvy_04$ended_at <-  mdy_hm(divvy_04$ended_at)
str(divvy_04)

```


```{r, warning=FALSE, results='hide'}
divvy_12$start_station_id <- as.numeric(as.character(divvy_12$start_station_id))
divvy_12$end_station_id <- as.numeric(as.character(divvy_12$end_station_id))
str(divvy_12)
```



Stacking all the monthly data frames and the quarterly data frame (divvy_q1) into one large data frame.

```{r}
all_rides <- bind_rows(divvy_q1, divvy_04, divvy_05, divvy_06, divvy_07, divvy_08,
                  divvy_09, divvy_10, divvy_11, divvy_12)

```

```{r, results='hide'}
glimpse(all_rides)
colnames(all_rides)
nrow(all_rides)
head(all_rides)
tail(all_rides)

```

Next is to sort the data frame by the earliest started day 

```{r}
all_rides <- arrange(all_rides, started_at)
head(all_rides)
tail(all_rides)

```

```{r, results='hide'}
dim(all_rides)
str(all_rides)
summary(all_rides)

```

Checking the user type consistency

```{r}
table(all_rides$member_casual)

```

Creating new columns for date, year, month, day and day_of_week 

```{r}
all_rides$date <- as.Date(all_rides$started_at)

```

```{r, results='hide'}
all_rides$month <- format(as.Date(all_rides$date), "%m")
all_rides$day <- format(as.Date(all_rides$date), "%d")
all_rides$year <- format(as.Date(all_rides$date), "%Y")
all_rides$day_of_week <- format(as.Date(all_rides$date), "%A")
head(all_rides)

```

Create a new column to get the duration of rides

```{r, results='hide'}
all_rides$ride_duration <- difftime(all_rides$ended_at,all_rides$started_at)
str(all_rides)
summary(all_rides)

```

Now convert the ride_duration column to numeric type

```{r}
all_rides$ride_duration <- as.numeric(as.character(all_rides$ride_duration))

```

```{r}
is.numeric(all_rides$ride_duration)

```

Remove any negative ride length in the ride_duration column

```{r}
all_rides_v2 <- all_rides[!(all_rides$ride_duration<0),]
summary(all_rides_v2)
summary(all_rides_v2$ride_duration)

```

#### DATA ANALYSIS
Compare members and casual users statistics:  
Mean

```{r}
aggregate(all_rides_v2$ride_duration ~ all_rides_v2$member_casual, FUN = mean)

```
Median
```{r}
aggregate(all_rides_v2$ride_duration ~ all_rides_v2$member_casual, FUN = median)

```
Maximum
```{r}
aggregate(all_rides_v2$ride_duration ~ all_rides_v2$member_casual, FUN = max)

```
Minimum
```{r}
aggregate(all_rides_v2$ride_duration ~ all_rides_v2$member_casual, FUN = min)

```
Total Durations
```{r}
aggregate(all_rides_v2$ride_duration ~ all_rides_v2$member_casual, FUN = sum)

```



See average ride time for each day by members and casual users

```{r results='hide'}
aggregate(all_rides_v2$ride_duration ~ all_rides_v2$member_casual + all_rides_v2$day_of_week,
          FUN = mean)

```
Noticing that the days of the week are out of order.

```{r}
all_rides_v2$day_of_week <- 
  ordered(all_rides_v2$day_of_week, levels=c("Sunday", "Monday",  "Tuesday", 
                                             "Wednesday", "Thursday", "Friday", "Saturday"))

```

```{r}
aggregate(all_rides_v2$ride_duration ~ all_rides_v2$member_casual + all_rides_v2$day_of_week,
          FUN = mean)

```

Analyze ridership data by type and weekday

```{r}
all_rides_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%    #creates weekday field using wday()
group_by(member_casual, weekday) %>%                      #groups by usertype and weekday
  summarise(number_of_rides = n(),        #calculates the number of rides and average duration
            average_duration = mean(ride_duration)) %>%     #calculates the average duration
  arrange(member_casual, weekday)                         #sorts

```

Visualizing the number of rides by rider type

```{r}
all_rides_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()
,average_duration = mean(ride_duration)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of rides per Weekday"
      )

```



Visualizing the average duration each weekday

```{r}
all_rides_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_duration)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average duration per Weekday"
      )

```


##### OBSERVATIONS
  
  * Annual members have the highest number of rides with a total value of 2,175,108.  
  * Casual users have the highest ride duration time with a total of  3,878,904,912 seconds.  
  * Annual members have the highest total number of daily rides with the exception of Saturday.  
  * Casual users have much more daily ride duration as compared to members.


I now saved the cleaned dataset file to local storage for further analysis and visualizations on Tableau.



##### TABLEAU  
Tableau Dashboard

#to embed tableau dashboard
#iframe("100%", 600, "https://public.tableau.com/views/CyclisticBike-Share2020/Dashboard1?:
#language=en-US&:display_count=n&:origin=viz_share_link:showVizHome=no&:embed=true")

[View on Tableau](https://public.tableau.com/app/profile/uko.etukho/viz/CyclisticBike-Share2020/Dashboard1)

##### OBSERVATIONS
  * Ride duration peak months range is from May to October with the highest month being July for both members and casual users.
  * Docked type of bike is most preferred type of bike used.
  * Casual users have the longest daily ride duration with peak days on Saturday and Sunday.
  * Members have the highest number of rides while Casual users have the highest duration of ride.
  
#### SUMMARY
Since peak usage period is between May and October as seen on the dashboard with a lot of activities by the casual users, a promotion package should be designed for intending members during this period to achieve maximum effect.  

Similarly, a package should be designed for intending members during the weekends to encourage casual users to migrate to membership as it is seen that most casual users ride more on weekends.  

Promotions should also be intensified within weekends and during peak months.
  
